---
title: "03 Normalization and PCA"
---

```{r}
#| include: false

# library
library(tidyverse)
library(Seurat)
library(data.table)
library(celldex)
library(SingleR)
library(ExperimentHub)
```

```{r}
#| include: false

filtered.seuratObj <- readRDS("data/intermediate_data/02_filtered.seuratObj.rds")
```

# Normalization

We will perform a normalization on UMI counts using `LogNormalize`.

`LogNormalize` divides the UMI counts of a gene in a cell into the total UMI counts in that cell. After division, we take the natural log.

```{r}
filtered.seuratObj <- NormalizeData(filtered.seuratObj, 
                                    normalization.method = "LogNormalize")
```

::: callout-important
## Question

Where are the normalized counts stored?
:::

::: {.callout-tip collapse="T"}
## Answer

The normalizaed counts are stored in the slot `RNA`.

```{r}
filtered.seuratObj@assays[["RNA"]]@data %>% head()
```
:::

# Evaluating effects of mitochondrial percentage

The unwanted variation, such as the difference in mitochondrial percentage, could affect the downstream analysis.

For example, we don't want the cells clustered based on the difference in mitochondrial percentage.

Let's first look at if the mitochondrial percentage introduces unwanted variation. We take the normalized counts and see if we have unwanted variation from the mitochondrial percentage.

1.  The mitochondrial percentage is a series of numbers. We can turn the mitochondrial percentage variable into a categorical variable based on quartiles.

```{r}
summary(filtered.seuratObj$percent.mt)
```

We can have three categories based on the quartiles:

Mitochondrial percentage below 1st quartile (10.2896%): Low Mitochondrial percentage between 1st and Median: Median Mitochondrial percentage between Median and 3rd quartile (15.4158%): Median high Mitochondrial percentage beyond 3rd quartile (15.4158%): High

Next, we create a new variable based on the cutoff.

```{r}
filtered.seuratObj$quartile.mt <- cut(filtered.seuratObj$percent.mt, 
                                      breaks=c(-Inf, 10.2856, 12.6322, 15.4088, Inf),
                                      labels=c("Low","Medium","Medium high", "High"))
```

We want to check whether the mitochondrial percentage is a source of variation using PCA. We plot the first two principal components to visualize the data.

```{r}
# Identify the most variable genes
filtered.seuratObj <- FindVariableFeatures(filtered.seuratObj,
                                           selection.method = "vst",
                                           nfeatures = 2000, 
                                           verbose = FALSE)
		     
# Scale the data
filtered.seuratObj <- ScaleData(filtered.seuratObj, features = rownames(filtered.seuratObj))

# perform PCA
filtered.seuratObj <- RunPCA(filtered.seuratObj)

# Plot pc1 and pc2
DimPlot(filtered.seuratObj,
        reduction = "pca",
        group.by= "quartile.mt",
        split.by = "quartile.mt")
```

Based on the above plot, we can see a different scatter pattern in cells with "Low" mitochondrial percentage.

We observe that the lobe of cells in the left middle side of the plot is in the category "Low". For all other levels of mitochondrial percentage, we see a more even distribution of cells across the PCA plot.

::: callout-tip
Oftentimes, it is helpful to regress out variation due to mitochondrial percentage. However, if the differences in mitochondrial gene expression represent a biological phenomenon that might help to distinguish cell clusters, then we do not regress this out.
:::

::: callout-tip
Cell cycle is another common variation. For instructions, please refer to <https://satijalab.org/seurat/articles/cell_cycle_vignette.html>
:::

To regress out the effect mitochondrial percentage, add the argument `vars.to.regress = "percent.mt"` to the `ScaleData`.

```{r}
# Scale the data
filtered.seuratObj <- ScaleData(filtered.seuratObj, 
                                features = rownames(filtered.seuratObj), 
                                vars.to.regress = "percent.mt")

# perform PCA
filtered.seuratObj <- RunPCA(filtered.seuratObj)
```
